Effectful Semantics for Substructural SSA

### 2023-11-13 pldi24/paper.tex
Tap

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we will to introduce a \textit{buffer} of write actions. We will
implement TSO ordering by buffering all our writes, in which case they are only
visible to the local thread. At any point, we may choose to \textit{flush} some
of the writes in buffer to the global pomset of events. In particular, we will
introduce the set \(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write
actions}, where an action \(\bufloc{x} := v\) by a thread denotes adding a write
\(x := v\) to the thread's write buffer. We may then define the set of TSO
actions \(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be
defined to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will
interpret as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index. In
particular, we define the monad \(\ms{TSO} =
\ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\), where \(\Sigma\) is the stream
action of finite pomsets over \(\mc{A}_{\ms{TSO}}\) on pomsets over
\(\mc{A}_{\ms{TSO}}\); we can view \(\ms{PO}\) as a submonad of this monad in
the standard manner.

Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
\begin{equation}
  (L;\{\bufloc{x} := v\})[x] = v
  \qquad
  (L;\{\bufloc{y} := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! 
\TODO{segue}
This is implemented via the morphism (called
\(\ms{split}\) in \cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}
\TODO{reflow}
On the other hand, reads events will first attempt to read from the buffer, and,
if there is no corresponding write in the buffer, will read an arbitrary value,
in both cases pushing an event to the global pomset; that is, the semantics of
reads and writes are given by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
  \end{aligned}
\end{equation}
\TODO{text explaining this}
\TODO{in the presence of parallel composition}

>>>>

To graduate from SC concurrency to a genuine, if maximally simple, weak memory
model, we will to introduce a \textit{buffer} of write actions. We will
implement TSO ordering by buffering all our writes, in which case they are only
visible to the local thread. On the other hand, reads events will first attempt
to read from the buffer, and, if there is no corresponding write in the buffer,
will read an arbitrary value, in both cases pushing an event to the global
pomset. At any point, we may choose to \textit{flush} some of the writes in
buffer to the global pomset of events. In particular, we will introduce the set
\(\mc{A}_b = \{(\bufloc{x} := v)\}\) of \textit{buffer write actions}, where an
action \(\bufloc{x} := v\) by a thread denotes adding a write \(x := v\) to the
thread's write buffer. We may then define the set of TSO actions
\(\mc{A}_{\ms{TSO}} = \mc{A}_{\ms{PO}} \cup \mc{A}_b\). A buffer will be defined
to be a list of write actions \(\ms{Buf} = \mc{A}_b^*\), which we will interpret
as linear pomsets over \(\mc{A}_{\ms{TSO}}\) ordered by index. In particular, we
define the monad \(\ms{TSO} = \ms{StateT}\;\ms{Buf}\;(\ms{Trace}\;\Sigma)\),
where \(\Sigma\) is the stream action of finite pomsets over
\(\mc{A}_{\ms{TSO}}\) on pomsets over \(\mc{A}_{\ms{TSO}}\); we can view
\(\ms{PO}\) as a submonad of this monad in the standard manner.

Given a buffer \(\ms{Buf}\), we can define the result of reads \([\cdot]_x:
\ms{Buf} \to \ms{Word} \sqcup \{\bot\}\) from the buffer as follows:
\begin{equation}
  (L;\{\bufloc{x} := v\})[x] = v
  \qquad
  (L;\{\bufloc{y} := v\})[x] = L[x]
  \qquad
  [][x] = \bot
\end{equation}
Note that writes at the \textit{end} of the buffer are prioritized, since later
writes overwrite earlier ones! The semantics of reads and writes are then given
by
\begin{equation}
  \begin{aligned}
  \dnt{\ms{read}_x} = R_x^{\ms{TSO}} 
    &= \ms{pflush};(\lambda ()\;L. \{(v, L, \{x = v\}) \mid L[x] = v \lor L[x] = \bot\});\ms{pflush} \\
  \dnt{\ms{write}_x} = W_x^{\ms{TSO}}
    &= \ms{pflush};(\lambda v\;L. \{(v, (L;\{\bufloc{x} := v\}), \{x := v\})\});\ms{pflush}
  \end{aligned}
\end{equation}
where buffer flushing is implemented via the morphism (called \(\ms{split}\) in
\cite{sparky})
\begin{equation}
  \ms{pflush} = \lambda ()\;L. \{((), R, \alpha) | L = \alpha;R\}
  : \ms{Set}_{\ms{TSO}}(\mb{1}, \mb{1})
\end{equation}
\TODO{text explaining this}
\TODO{in the presence of parallel composition}
